---
title: "Simulation experiment (5)"
format: 
  html:
    embed-resources: true
toc: true
---

```{r packages, echo = F}
library(tidyverse) |> suppressPackageStartupMessages()
library(flextable) |> suppressPackageStartupMessages()
library(targets) |> suppressPackageStartupMessages()
library(tarchetypes) |> suppressPackageStartupMessages()
library(here) |> suppressPackageStartupMessages()

# set the _target store library of the project (for interactive use)
prj_store <- here("store_simexp5")
tar_load(predeval_summary, store = prj_store)
nrep <- length(unique(predeval_summary$.rep)) 
```


## Design of the experiment

The design of the experiment is analogous to the experiment with normally distributed data, but some alterations were made to accommodate strictly postive variables (eg. ECEC, bulk density, ...).

We simulate data according to the following method (notation: $X$ = non-reference method, $Y$ = reference method). We first simulate a latent (true) value for the non-reference method using a Gamma distribution with shape parameter $\alpha$ and rate parameter $\beta$. The Gamma distribution parameters are related to the mean and coefficient of variation ($CV = \sigma/\mu$) of $X$ and we specify those to create a scenario:

$$
\begin{align}
\alpha &= CV_X^{-2} \\
\beta &=  \frac{CV_X^{-2}}{\mu_X} \\
X_{\mathrm{true}} &\sim \mathcal{G}(\alpha, \beta)
\end{align}
$$

We then use this value to simulate a value for the latent reference method $Y$, assuming a linear relation between both methods on the log-scale and some structural noise that is not due to measurement error. This structural noise needs to be multiplicative to avoid negative values; we sample $Y$ therefore from a Gamma distribution with a shape parameter reflecting the residual coefficient of variation (conditional on $X$; $CV_{Y|X}$):

$$
\begin{align}
\mu_Y &= \exp\left(\gamma_0 + \gamma_1\log(X_{\mathrm{true}})\right) \\
Y_{\mathrm{true}} &\sim \mathcal{G}\left(CV_{Y|X}^{-2}, \frac{CV_{Y|X}^{-2}}{\mu_Y}\right)
\end{align}
$$

We then add measurement noise to both methods (resulting from both field and lab measurement processes). Again, the noise is multiplicative ($X_{\mathrm{obs}} = X_{\mathrm{true}}\epsilon_u$, $Y_{\mathrm{obs}} = Y_{\mathrm{true}}\epsilon_v$) to avoid negative values and we take Gamma-distributed noises with expected values 1 (unbiased noise). Just like in the gaussian case, the measurement noise magnitudes are expressed as fractions of the $CV$ of the latent $X$:

$$
\begin{aligned}
CV_{\epsilon_u} &= \tau_xCV_X \\
CV_{\epsilon_v} &= \tau_{xy}CV_{\epsilon_u}
\end{aligned}
$$

We allow for correlated measurement noise between both methods using a Gaussian copula; we start by sampling a bivariate variable $Z$ with zero mean, unit variance and correlation coefficient $\rho$, this variable is then transformed to the unit-interval by means of the standard-normal CDF $\Phi_\mathcal{N}$ and then transformed to a Gamma-distributed measurement error by means of the Gamma quantile function $\Phi_\mathcal{G}$:

$$
\begin{align}
\begin{pmatrix} Z_1\\ Z_2 \end{pmatrix} 
&\sim \mathcal{N}
\left(
\begin{pmatrix} 0\\ 0 \end{pmatrix},
\begin{pmatrix} 1 & \rho \\ \rho & 1 \end{pmatrix}
\right) \\
U_1 &= \Phi_\mathcal{N}(Z_1) \\
U_2 &= \Phi_\mathcal{N}(Z_2) \\
\epsilon_u &= \Phi_\mathcal{G}^{-1}(CV_{\epsilon_u}^{-2}, CV_{\epsilon_u}^{-2}) \\
\epsilon_v &= \Phi_\mathcal{G}^{-1}(CV_{\epsilon_v}^{-2}, CV_{\epsilon_v}^{-2})
\end{align}
$$



## Models

### Linear regression

### Linear regression using log-transformed values

### Gamma GLM

### Gamma EIV-model with known CV of the measurement error

### Gamma EIV-model with unknown CV of the measurement error


<br>

## Results

### Predictions

### Model diagnostics: MCMC $\hat{R}$

The following figures display the $\hat{R}$ statistic of the MCMC for each of the model parameters. The red dashed lines are at 1.00 and 1.01. The boxplots are aggregated across scenario's. The Gamma EIV-model with unknown $CV$ of the measurement error suffers from poor mixing.

```{r mcmc-dx, echo = F, warning = F}
tar_load(mcmcdx_summary,  store = prj_store)

mcmcdx_summary %>%
  select(-c(sample_size, mux, cvx, alpha, beta, 
            cvy, taux, tauxy, corr_mexy)) %>%
  pivot_longer(-c(.rep, model), names_to = "parameter", values_to = "Rhat") %>%
  filter(!is.na(Rhat)) %>%
  mutate(
    model = factor(
      model, c("linreg","linreglogtrafo","gammareg",
               "gammaregeivknowncvmex","gammaregeivunknowncvmex"),
      c("Linear regression","Linear regression\n(logtransformed data)",
        "Gamma GLM","Gamma EIV\n(known CV mex)","Gamma EIV\n(prior on CV mex)")
    )
  ) %>%
  ggplot(aes(Rhat, parameter)) +
  facet_wrap(~ model) +
  geom_boxplot() +
  geom_vline(xintercept = c(1.00, 1.01), lty = 2, color = "red") +
  theme_minimal()
```

Looking at the MCMC samples from 1 run of this model also reveals the same issue as in the gaussian EIV-model: high negative correlated parameters.

```{r, warning=F}
tar_load(mcmc_gammareg_eiv_unknowncvmex_Nsample200_Mux1_CVx0.5_Alpha0_Beta1_CVy0.01_Taux0.05_Tauxy1_Rho0_162fe9cbcc197974, store = prj_store)
mcmc_gammareg_eiv_unknowncvmex_Nsample200_Mux1_CVx0.5_Alpha0_Beta1_CVy0.01_Taux0.05_Tauxy1_Rho0_162fe9cbcc197974 %>%
   filter(.rep == "e99fbf80fca2d92e") %>%
   rename(chain = .chain) %>%
   select(!contains("[")) %>% select(!contains(".")) %>% 
   select(-c(lp__, shape_x, rate_x)) %>%
   GGally::ggscatmat(color = "chain", corMethod = "spearman")
```


