---
title: "Simulation experiment (8)"
format: 
  html:
    embed-resources: true
toc: true
---

# Design of the experiment

This experiment is designed based on actual soil data from EJP Soil. Below we show ECEC (mmol/kg) as measured by the central Eurofins laboratory (using ISO-23470) vs. the measurements from national labs (where available). Belgium-Wallonia and France used the same method as Eurofins, while other countries used different methods and require a transfer function.

```{r, echo = F, message=F, warning=F}
library(tidyverse) |> suppressPackageStartupMessages()

load(paste0("C:/Users/ben_serrien/Documents/SoilHarmony/",
            "soildata-Eurofins-FlandersDS/data/data_ecec.rda"))

ggplot(data_ecec, aes(ecec_nsims, ecec_eurofins)) +
  facet_wrap(~ country + method_nsims) +
  geom_point() +
  geom_smooth(se = F) +
  geom_smooth(se = F, method = "lm", color = "red") +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  labs(y = "ECEC (Eurofins: ISO-23470)",
       x = "ECEC (NSIMS: non-reference method)")
```


## Austria

Let's take Austria as an example. Below we show the data and some summary statistics of the ECEC with the NSIMS method.

```{r, echo = F, message=F, warning=F}
ecec_austria <- data_ecec %>%
  filter(country == "Austria")

ggplot(ecec_austria, aes(ecec_nsims, ecec_eurofins)) +
  geom_point() +
  geom_smooth(se = F) +
  geom_smooth(se = F, method = "lm", color = "red") +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  xlim(c(0,NA)) + ylim(c(0,NA)) +
  labs(y = "ECEC (Eurofins: ISO-23470)",
       x = "ECEC (NSIMS: ONORM L 1086-1)")

ecec_austria %>%
  summarise(
    mean_ecec_nsims = mean(ecec_nsims),
    sd_ecec_nsims = sd(ecec_nsims),
    cv_ecec_nsims = sd_ecec_nsims/mean_ecec_nsims
  )
```

The idea is to design a simulation experiment as follows. We fit the Gamma EIV-model to these data. The purpose of this model is to use it to simulate new data-sets that are similar to the Austrian ECEC data but whose properties we know exactly so we can calculate bias and coverage metrics.

```{r}
library(cmdstanr) |> suppressPackageStartupMessages()

mod <- cmdstan_model(here::here("source/stan/gammareg_eiv_AustriaECEC.stan"),
                     compile = TRUE)

stan_list <- list(
  N     = nrow(ecec_austria),
  x_obs = as.numeric(ecec_austria$ecec_nsims),
  y_obs = as.numeric(ecec_austria$ecec_eurofins)
)

fit <- mod$sample(data = stan_list, adapt_delta = 0.80,
                  show_messages = T, show_exceptions = T)

# summary statistics
fit$summary(
  variables = c("mu_x","cv_x","shape_mex","beta0","beta0c","beta1","shape")
)

draws_params <- fit$draws(
  variables = c("mu_x","cv_x","shape_mex","beta0","beta0c","beta1","shape"),
  format = "draws_df"
)
draws_params %>%
  select(-c(.draw, .iteration)) %>%
  GGally::ggscatmat(color = ".chain")

# goodness-of-fit
y_rep <- fit$draws("y_rep", format = "matrix")
bayesplot::ppc_dens_overlay(y = stan_list$y_obs, yrep = y_rep[1:100, ])
```

We now have 



```{r}

draws_params <- fit$draws(
  variables = c("mu_x","cv_x","shape_x","shape_mex",
                "beta0","beta0c","beta1","shape"),
  format = "draws_df"
)

source(here::here("source/R/simulate-data-gamma.R"))

data_ecec_sim <- sim_data_gamma(
  N    = 200, 
  mu_x = draws_params$mu_x[1], 
  cv_x = draws_params$cv_x[1],
  cv_y = 1/sqrt(draws_params$shape[1]),
  alpha = draws_params$beta0[1],
  beta  = draws_params$beta1[1],
  ratio_cvmex_cvx = .05,
  #ratio_cvmex_cvx = (1/sqrt(draws_params$shape_mex[1])) / draws_params$cv_x[1],
  ratio_mey_mex = 1,
  corr_error    = 0
)

data_ecec_sim_cal <- data.frame(
  x_obs = data_ecec_sim$x_obs,
  y_obs = data_ecec_sim$y_obs,
  x_true = data_ecec_sim$x_true,
  y_true = data_ecec_sim$y_true
)

ggplot(data_ecec_sim_cal) +
  geom_point(aes(x_obs, y_obs)) +
  geom_point(aes(x_true, y_true), color = "red") +
  geom_abline(slope = 1, intercept = 0, lty = 2)
```



