---
title: "Simulation experiment (5bis)"
format: 
  html:
    embed-resources: true
toc: true
---

```{r packages, echo = F}
library(tidyverse) |> suppressPackageStartupMessages()
library(flextable) |> suppressPackageStartupMessages()
library(targets) |> suppressPackageStartupMessages()
library(tarchetypes) |> suppressPackageStartupMessages()
library(here) |> suppressPackageStartupMessages()

# set the _target store library of the project (for interactive use)
prj_store <- here("store_simexp5bis")
```


## Design of the experiment

Analogous to experiment 5. In this experiment we focus on the effect of the measurement error structure ($\tau_x, \tau_{xy}, \rho$). Other parameter values are kept fixed:

```{r, echo=F}
tar_load(predeval_summary, store = prj_store)
nrep <- length(unique(predeval_summary$.rep)) 
```

| Parameter              | Values                                      |
|------------------------|---------------------------------------------|
|$N$                     |`r unique(predeval_summary$sample_size)`     |
|$\mu_X$                 |`r unique(predeval_summary$mux)`             |
|$CV_X$                  |`r unique(predeval_summary$cvx)`             |
|$\alpha$                |`r unique(predeval_summary$alpha)`           |
|$\beta$                 |`r unique(predeval_summary$beta)`            |
|$CV_{Y|X}$              |`r unique(predeval_summary$cvy)`             |
|$\tau_x$                |`r unique(predeval_summary$taux)`            |
|$\tau_{xy}$             |`r unique(predeval_summary$tauxy)`           |
|$\rho$                  |`r unique(predeval_summary$corr_mexy)`       |

: Table: Parameter values for the scenario's of experiment 5bis.

<br>

## Models

Same as in experiment 5.

<br>

## Results

### Predictions

Figure 

```{r, echo=F}
#| label: fig-calibration1
#| fig-cap: "Results of the simulation experiment: effect of the intercept of the XY-relation (log-scale) on calibration metrics (aggregated across different levels of other experimental parameters). Each facet has a different y-axis because the calibration metrics are on different scales."

predeval_summary %>%
  pivot_longer(
    c(PICP, MAE, MPE, RMSE, R2, SDPE), 
    names_to = "metric", values_to = "value"
  ) %>% 
  mutate(
    metric = factor(metric, c("MPE","MAE","RMSE","R2","SDPE","PICP")),
    model = factor(
      model, c("linreg","linreglogtrafo","gammareg",
               "gammaregeivknowncvmex","gammaregeivunknowncvmex"),
      c("Linear regression","Linear regression\n(logtransformed data)",
        "Gamma GLM","Gamma EIV\n(known CV mex)","Gamma EIV\n(prior on CV mex)")
    )
  ) %>%
  ggplot(aes(factor(alpha), value, colour = model)) +
  facet_wrap(~ metric, scales = "free_y") +
  geom_boxplot() +
  geom_point(position = position_jitterdodge(.1, 0)) +
  labs(
    title = "calibration metrics", y = "",
    x = "intercept of the XY-relation on the log-scale",
    color = "Models"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "bottom")
```

```{r, echo=F}
#| label: fig-calibration2
#| fig-cap: "Results of the simulation experiment: effect of the slope of the XY-relation (log-scale) on calibration metrics (aggregated across different levels of other experimental parameters). Each facet has a different y-axis because the calibration metrics are on different scales."

predeval_summary %>%
  pivot_longer(
    c(PICP, MAE, MPE, RMSE, R2, SDPE), 
    names_to = "metric", values_to = "value"
  ) %>%
  mutate(
    metric = factor(metric, c("MPE","MAE","RMSE","R2","SDPE","PICP")),
    model = factor(
      model, c("linreg","linreglogtrafo","gammareg",
               "gammaregeivknowncvmex","gammaregeivunknowncvmex"),
      c("Linear regression","Linear regression\n(logtransformed data)",
        "Gamma GLM","Gamma EIV\n(known CV mex)","Gamma EIV\n(prior on CV mex)")
    )
  ) %>%
  ggplot(aes(factor(beta), value, colour = model)) +
  facet_wrap(~ metric, scales = "free_y") +
  geom_boxplot() +
  geom_point(position = position_jitterdodge(.1, 0)) +
  labs(
    title = "calibration metrics", y = "",
    x = "slope of the XY-relation on the log-scale",
    color = "Models"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "bottom")
```

```{r, echo=F}
#| label: fig-calibration3
#| fig-cap: "Results of the simulation experiment: effect of the slope and intercept of the XY-relation (log-scale) on PICP. Each facet has a different y-axis because the calibration metrics are on different scales."

predeval_summary %>%
  mutate(
    model = factor(
      model, c("linreg","linreglogtrafo","gammareg",
               "gammaregeivknowncvmex","gammaregeivunknowncvmex"),
      c("Linear regression","Linear regression\n(logtransformed data)",
        "Gamma GLM","Gamma EIV\n(known CV mex)","Gamma EIV\n(prior on CV mex)")
    )
  ) %>%
  ggplot(aes(model, PICP, colour = model)) +
  facet_grid(alpha ~ beta, labeller = "label_both") +
  geom_boxplot() +
  geom_hline(yintercept = .95, lty = 2) +
  geom_point(position = position_jitterdodge(.1, 0)) +
  labs(
    title = "calibration metrics", y = "PICP", x = "",
    color = "Models"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        axis.text.x = element_blank(),
        legend.position = "bottom")
```

```{r, echo=F}
#| label: fig-calibration4
#| fig-cap: "Results of the simulation experiment: effect of the slope and intercept of the XY-relation (log-scale) on R². Each facet has a different y-axis because the calibration metrics are on different scales."

predeval_summary %>%
  mutate(
    model = factor(
      model, c("linreg","linreglogtrafo","gammareg",
               "gammaregeivknowncvmex","gammaregeivunknowncvmex"),
      c("Linear regression","Linear regression\n(logtransformed data)",
        "Gamma GLM","Gamma EIV\n(known CV mex)","Gamma EIV\n(prior on CV mex)")
    )
  ) %>%
  ggplot(aes(model, R2, colour = model)) +
  facet_grid(alpha ~ beta, labeller = "label_both") +
  geom_boxplot() +
  geom_point(position = position_jitterdodge(.1, 0)) +
  labs(
    title = "calibration metrics", y = "R²", x = "",
    color = "Models"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        axis.text.x = element_blank(),
        legend.position = "bottom")
```

The linear regression model suffers from the inability to constrain predictions and prediction intervals to the positive real line. The other models don't have this deficiency.

```{r, echo=F}
predeval_summary %>%
  filter(model != "linreg") %>%
  select(model, PNEG, PNEGll) %>%
  summarise(
    .by = model, 
    `data points with at least 1 positive prediction` = sum(PNEG > 0),
    `data points with at least 1 positive prediction interval` = sum(PNEGll > 0)
  )
```


```{r, warning=F, echo=F}
#| label: fig-calibration5
#| fig-cap: "Linear regression: proportion of data points with negative predictions or negative prediction intervals"

predeval_summary %>%
  filter(model == "linreg") %>%
  pivot_longer(
    c(PNEG, PNEGll), 
    names_to = "metric", values_to = "value"
  ) %>% 
  mutate(metric = factor(metric, c("PNEG","PNEGll"))) %>%
  ggplot(aes(metric, 100*value)) +
  geom_boxplot() +
  scale_y_log10() +
  geom_point(position = position_jitterdodge(.1, 0)) +
  labs(
    title = "calibration metrics", 
    y = "percentage data points with negative values",
    x = "metric"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "bottom")
```

```{r, warning=F, echo=F}
#| label: fig-calibration6
#| fig-cap: "Results of the simulation experiment: %bias and %variance"

predeval_summary %>%
  mutate(
    pct_bias = MPE^2 / RMSE^2,
    model = factor(
      model, c("linreg","linreglogtrafo","gammareg",
               "gammaregeivknowncvmex","gammaregeivunknowncvmex"),
      c("Linear regression","Linear regression\n(logtransformed data)",
        "Gamma GLM","Gamma EIV\n(known CV mex)","Gamma EIV\n(prior on CV mex)")
    )
  ) %>%
  ggplot(aes(factor(alpha), pct_bias, color = model)) +
  geom_boxplot(position = position_dodge(.8)) +
  geom_jitter(position = position_dodge(.8)) +
  facet_wrap(~ beta, labeller = "label_both") +
  scale_y_continuous(
    name = "% bias", labels = scales::percent,
    sec.axis = sec_axis(~ 1 - ., name = "% variance", labels = scales::percent)
  ) +
  labs(x = "alpha") +
  theme_minimal()
```

```{r, warning=F, echo=F}
#| label: fig-calibration7
#| fig-cap: "Results of the simulation experiment: MPE vs. SDPE. All points on the circles have the same RMSE. Aggregated across scenario's"

predeval_summary %>%
  mutate(
    model = factor(
      model, c("linreg","linreglogtrafo","gammareg",
               "gammaregeivknowncvmex","gammaregeivunknowncvmex"),
      c("Linear regression","Linear regression\n(logtransformed data)",
        "Gamma GLM","Gamma EIV\n(known CV mex)","Gamma EIV\n(prior on CV mex)")
    )
  ) %>%
  ggplot(aes(MPE, SDPE, color = model)) +
  geom_point(alpha = .3) +
  ggforce::geom_circle(inherit.aes = F, aes(x0=0, y0=0, r=.1)) +
  ggforce::geom_circle(inherit.aes = F, aes(x0=0, y0=0, r=.25)) +
  coord_fixed(ylim = c(0, NA), 
              xlim = c(1.1*min(predeval_summary$MPE), 
                       1.1*max(predeval_summary$MPE))) +
  geom_vline(xintercept = 0, lty = 2) +
  geom_hline(yintercept = 0, lty = 2) +
  theme_minimal()
```

### Model diagnostics: MCMC $\hat{R}$

The following figures display the $\hat{R}$ statistic of the MCMC for each of the model parameters. The red dashed lines are at 1.00 and 1.01. The boxplots are aggregated across scenario's. The Gamma EIV-model with unknown $CV$ of the measurement error suffers from poor mixing.

```{r mcmc-dx, echo = F, warning = F}
tar_load(mcmcdx_summary,  store = prj_store)

mcmcdx_summary %>%
  select(-c(sample_size, mux, cvx, alpha, beta, 
            cvy, taux, tauxy, corr_mexy)) %>%
  pivot_longer(-c(.rep, model), names_to = "parameter", values_to = "Rhat") %>%
  filter(!is.na(Rhat)) %>%
  mutate(
    model = factor(
      model, c("linreg","linreglogtrafo","gammareg",
               "gammaregeivknowncvmex","gammaregeivunknowncvmex"),
      c("Linear regression","Linear regression\n(logtransformed data)",
        "Gamma GLM","Gamma EIV\n(known CV mex)","Gamma EIV\n(prior on CV mex)")
    )
  ) %>%
  ggplot(aes(Rhat, parameter)) +
  facet_wrap(~ model, scales = "free_x") +
  geom_boxplot() +
  geom_vline(xintercept = c(1.00, 1.01), lty = 2, color = "red") +
  theme_minimal()
```
